---
title: "Industrial software dataset"
author: "Eva Ssozi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-industrial dataset from the internet, source:https://datausa.io/profile/soc/software-developers-applications-systems-software
```{r}
#libraries required:
library(tidyverse)
library(readxl)
library(readr)
library(esquisse)

library(cluster)
library(fpc)
library(clustMixType)
library(NbClust)
```

# **Exploratory Data Analysis, including K-means cluster analysis**
Showcasing EDA process, including K-means clustering.

```{r}
#Import dataset
Industries_by_Share <- read_csv("Industries by Share.csv")
View(Industries_by_Share)
```


```{r}
#On observation, this dataset has a number of columns that are all one value. Investigate to eliminate these.
#change character types to factor type:
df <- Industries_by_Share %>%
  mutate_if(is.character, as.factor)
```


```{r}
#view summaries for all variables
summary(df)
```

```{r}
#frequency tables for all the factor variables in df: to see which variables have only one level and can therefore be excluded from further analysis

#select only factor variables
df_factor_freq <- df %>%
  select_if(is.factor)

#calculate frequency
apply(df_factor_freq, 2, table)
```

```{r}
#exclude columns with only one level from df:
#ID columns are directly paired with the columns they represent, so they are also excluded
df_2 <- df %>%
  select(Industry_Sector,Industry_Sub_Sector,Industry_Group,Year,Workforce_Status,Total_Population,Total_Population_MOE_Appx,Average_Wage,Average_Wage_Appx_MOE,Record_Count,Slug_Industry_Group)
```

```{r}
#checking workforce status column, seems to be filled with just "TRUE":
check_workforce <- df_2 %>%
  filter(Workforce_Status == FALSE)

check_workforce #empty, so remove this column as well

df_2 <- df_2 %>% select(-Workforce_Status)
```

```{r}
#checking for nas in the columns of df_2:
apply(df_2, 2, function(x) sum(is.na(x)))
#column Slug_Industry_Group has 8 rows with NA
```

```{r}
#inspect the NA's found:
check_na <- df_2 %>% filter(is.na(Slug_Industry_Group))
check_na #two sectors: construction and Professional(4); Scientific & Management, and Administrative & Waste Management Services(4 out of many)

#exclude these nas:
df_2 <- na.omit(df_2)
```

Bar chats for some of the categorical variables' population: Plotting population and Average wage for the industry sectors and sub-sectors.

```{r}
#change year to factor for visualisation prep:
df_2$Year <- as.factor(df_2$Year)
levels(df_2$Year) #2014, 2015, 2016, 2017
```

- Population for Industry Sector
```{r}
indust_sector <- df_2 %>%
  ggplot(aes(x = Total_Population, y = Industry_Sector)) +
  facet_grid(.~ Year) +
  geom_bar(stat = "identity") +
  theme(legend.position = "bottom") +
  labs(x = "Population", y = "Industry Sector")

indust_sector
```

- Population for Industry sub Sector
```{r}
indust_sub_sector <- df_2 %>%
  ggplot(aes(x = Total_Population, y = Industry_Sub_Sector)) +
  facet_grid(.~ Year) +
  geom_bar(stat = "identity") +
  theme(legend.position = "bottom") +
  labs(x = "Population", y = "Industry sub sector")

indust_sub_sector
```

Bar chats for some of the categorical variables' average wages:

- Average Wages for Industry Sector
```{r}
indust_sector <- df_2 %>%
  ggplot(aes(x = Average_Wage, y = Industry_Sector)) +
  facet_grid(.~ Year) +
  geom_bar(stat = "identity") +
  theme(legend.position = "bottom") +
  labs(x = "Average Wage", y = "Industry Sector")

indust_sector
```


- Average Wages for Industry Sub-Sector
```{r}
indust_sub_sector <- df_2 %>%
  ggplot(aes(x = Average_Wage, y = Industry_Sub_Sector)) +
  facet_grid(.~ Year) +
  geom_bar(stat = "identity") +
  theme(legend.position = "bottom") +
  labs(x = "Average Wage", y = "Industry sub sector")

indust_sub_sector
```

-Individual plots for each Year: Variation of population, average wage and record count by year:
```{r}
#data prep
df_3 <- df_2 %>%
  mutate(totalwage = Average_Wage * Total_Population) %>%
  group_by(Industry_Sector,Industry_Sub_Sector,Year) %>%
  summarise(allpopn = sum(Total_Population),
            groupwage = sum(totalwage)) %>%
  mutate(averagewage_subsector_year = groupwage/allpopn)
```

*2014, population:*

```{r}
indust_sub_sector_2014_popn <- df_3 %>%
  ungroup() %>%
  filter(Year == "2014") %>%
  ggplot(aes(x = allpopn, y = reorder(Industry_Sub_Sector, -allpopn))) +
  geom_bar(stat = "identity") +
  theme(legend.position = "bottom") +
  labs(x = "Population", y = "Industry sub sector")

indust_sub_sector_2014_popn
```


*All years, population, sub-sectors:*
```{r}
indust_sub_sector_popn <- df_3 %>%
  ungroup() %>%
  ggplot(aes(x = allpopn, y = reorder(Industry_Sub_Sector, -allpopn), fill = Year)) +
  geom_bar(stat = "identity") +
  facet_grid(.~Year) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  labs(x = "Population", y = "Industry sub sector")

indust_sub_sector_popn
```


*All years, average wage, sub-sectors:*
```{r}
indust_sub_sector_avwage <- df_3 %>%
  ungroup() %>%
  ggplot(aes(x = averagewage_subsector_year, y = reorder(Industry_Sub_Sector, -averagewage_subsector_year), fill = Year)) +
  geom_bar(stat = "identity") +
  facet_grid(.~Year) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  labs(x = "Average wage", y = "Industry sub sector")

indust_sub_sector_avwage
```


*cluster analysis by year: using population and average wage for each sub-sector*
```{r}
#extract columns for clustering
cluster_df <- df_3 %>%
  ungroup() %>%
  select(Industry_Sub_Sector,Year,allpopn,averagewage_subsector_year)

cluster_df
```


```{r}
#cluster preparation: standardisation only, since there are only 2 columns for clustering, no dimension reduction is required.
cluster_df[,3:4] <- scale(cluster_df[,3:4])
cluster_df

```


*Year = 2014*
```{r}
#data prep for cluster input:
df_2014 <- cluster_df %>%
  filter(Year == "2014") %>%
  select(allpopn,averagewage_subsector_year)

df_2014
#dimensions:
dim(df_2014) #18, 2
```

*Only 18 observations, so decided to limit number of clusters search to 6*

a) Number of clusters using elbow method: elbow at 2 or 3

```{r}
#Elbow method for k-means clustering, raw data*

set.seed(1)

#compute and plot wss for k =2 to k =6
k.max <- 6 #maximum number of clusters

wss <- sapply(1:k.max, function(k){kmeans(df_2014, k, nstart = 25)$tot.withinss}) #sapply(x, function(x))

plot(1:k.max, wss, type="b", pch=19, frame=T, xlab="Number of clusters K", ylab="Total within-clusters sum of squares", main="Total Within Clusters sum of squares for Kmeans")
```


b) Silhouette method for choosing k for k-means clustering: suggests 2 clusters
```{r}
#computing euclidean pairwise distances
df_2014.dist <- dist(df_2014, method="euclidean")

k.max <- 6

sil <- rep(0, k.max)

#compute the average silhouette width for k=2 to k=k.max
for(i in 2:k.max){
  km.res <- kmeans(df_2014, centers=i, nstart=25) #kmeans with i clusters (from 2 to k.max clusters)
  ss <- silhouette(km.res$cluster,df_2014.dist) #calculating silhouette widths
  sil[i] <- mean(ss[,3]) #average of silhouette widths which are in column 3
}

#plot the average silhouette width
plot(1:k.max, sil, type="b", pch=19, frame=T, xlab="Number of clusters k", ylab = "Average Silhouette Width", main = "Average Silhouette Widths for Kmeans")
abline(v=which.max(sil), lty=2)
```

c) NBClust method: suggests 3 clusters
```{r}
res.nb1 <- NbClust(df_2014, distance = "euclidean", min.nc = 2, max.nc = 6, method = "kmeans", index = "all")
#best number of clusters is 3
```


k means: using 3 clusters
```{r}
set.seed(1)
kmeans_df2014 <- kmeans(df_2014, 3, nstart = 25)
kmeans_df2014
#cluster sizes are 11,6 and 1
```

Extracting silhouette widths of the clusters
```{r}
y <- silhouette(kmeans_df2014$cluster, df_2014.dist)
#plot(y)

sil_info <- summary(y)
sil_info$clus.sizes
sil_info$clus.avg.widths
sil_info$avg.width #0.567, reasonable structure

#investigating between clusters sum of squares
(kmeans_df2014$betweenss)/(kmeans_df2014$totss)
```


```{r}
#add cluster assignments to original data (2014):
cluster_df_2014 <- df_3 %>%
  ungroup() %>%
  select(Industry_Sub_Sector,Year,allpopn,averagewage_subsector_year) %>%
  filter(Year == "2014")

cluster_df_2014$cluster <- kmeans_df2014$cluster
cluster_df_2014
```


Plot and colour by cluster:
```{r}
cluster_df_2014$cluster <- as.factor(cluster_df_2014$cluster)
```

```{r}
#cluster plot:
cluster_2014_plot <- cluster_df_2014 %>%
  ggplot(aes(x = allpopn, y = averagewage_subsector_year, label = Industry_Sub_Sector)) +
  geom_point(aes(color = cluster), size = 3) +
  geom_text(hjust = 0, vjust = 0)
  theme(legend.position = "right") +
  labs(x = "Population", y = "Average Wage")

cluster_2014_plot
```

*Year = 2017*
```{r}
#data prep for cluster input:
df_2017 <- cluster_df %>%
  filter(Year == "2017") %>%
  select(allpopn,averagewage_subsector_year)

df_2017

dim(df_2017) #19, 2
```

*Only 19 observations, so decided to limit number of clusters search to 6*

a) Number of clusters using elbow method: elbow at 3 or 4
```{r}
#Elbow method for k-means clustering, raw data*

set.seed(1)

#compute and plot wss for k =2 to k =6
k.max <- 6 #maximum number of clusters

wss <- sapply(1:k.max, function(k){kmeans(df_2017, k, nstart = 25)$tot.withinss}) #sapply(x, function(x))

plot(1:k.max, wss, type="b", pch=19, frame=T, xlab="Number of clusters K", ylab="Total within-clusters sum of squares", main="Total Within Clusters sum of squares for Kmeans")
```


b) Silhouette method for choosing k for k-means clustering: suggests 5 clusters
```{r}

#computing euclidean pairwise distances
df_2017.dist <- dist(df_2017, method="euclidean")

k.max <- 6

sil <- rep(0, k.max)

#compute the average silhouette width for k=2 to k=k.max
for(i in 2:k.max){
  km.res <- kmeans(df_2017, centers=i, nstart=25) #kmeans with i clusters (from 2 to k.max clusters)
  ss <- silhouette(km.res$cluster,df_2017.dist) #calculating silhouette widths
  sil[i] <- mean(ss[,3]) #average of silhouette widths which are in column 3
}

#plot the average silhouette width
plot(1:k.max, sil, type="b", pch=19, frame=T, xlab="Number of clusters k", ylab = "Average Silhouette Width", main = "Average Silhouette Widths for Kmeans")
abline(v=which.max(sil), lty=2)
```

c) NBClust method: suggests 4 clusters
```{r}
res.nb1 <- NbClust(df_2017, distance = "euclidean", min.nc = 2, max.nc = 6, method = "kmeans", index = "all")
#best number of clusters is 4
```


k means clustering:
```{r}
#k means clustering:
set.seed(1)
kmeans_df2017 <- kmeans(df_2017, 4, nstart = 25)
kmeans_df2017
#cluster sizes are 6,1,2,10
```

Extracting silhouette widths of the clusters
```{r}
y <- silhouette(kmeans_df2017$cluster, df_2017.dist)
#plot(y)

sil_info <- summary(y)
sil_info$clus.sizes
sil_info$clus.avg.widths
sil_info$avg.width #0.504, indicating a reasonable structure

#investigating between clusters sum of squares
(kmeans_df2017$betweenss)/(kmeans_df2017$totss)
```


```{r}
#add cluster assignments to original data (2014):

cluster_df_2017 <- df_3 %>%
  ungroup() %>%
  select(Industry_Sub_Sector,Year,allpopn,averagewage_subsector_year) %>%
  filter(Year == "2017")

cluster_df_2017$cluster <- kmeans_df2017$cluster
cluster_df_2017
```
Plot and colour by cluster:
```{r}
cluster_df_2017$cluster <- as.factor(cluster_df_2017$cluster)
```

```{r}
cluster_2017_plot <- cluster_df_2017 %>%
  ggplot(aes(x = allpopn, y = averagewage_subsector_year, label = Industry_Sub_Sector)) +
  geom_point(aes(color = cluster), size = 3) +
  geom_text(hjust = 0, vjust = 0)
  theme(legend.position = "right") +
  labs(x = "Population", y = "Average Wage")

cluster_2017_plot
```

**In both 2014 and 2017, "Professional, Scientific & Technical Services" was in a cluster of its own, could be an outlier**
**This sub sector has the highest population, but the average wage is closer to the middle when compared to the other sectors**


Testing github pulling and pushing requests:
-this is 2nd push to github
